Concurrent Programming with CXL

### 2020-05-18 book/paper.tex
Working on book...

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The $x..y$ notation generates a set containing the numbers
from~$x$ to~$y$ (inclusive).  The last two lines in the program are
the most interesting.
They assign to $x$ some unspecified value in the range $0..N$, and then
verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.
If the verification fails, the value of $x$ is reported.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def process(self):
   0 Jump 12
   1 Frame ('process', 'Up.cxl', 1, 5) [('self', '<stdin>', 1, 13)] 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
==== Safety violation ====
__main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process/0   [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process/1   [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
process/0   [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
__main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{dict\{ .name: .\_\_main\_\_, .tag: () \}}, pc 0,
register \texttt{()}, and an empty stack.
(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

\chapter{Using CXL Methods}

\chapter{Spinlock}

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.

>>>>

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
They assign to \texttt{x} some unspecified value in the range \texttt{0..N},
and then verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def process(self):
   0 Jump 12
   1 Frame ('process', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 13)] 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
==== Safety violation ====
__main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process /0  [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process /1  [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
process /0  [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
__main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

A name tag consists of the name of the main method of the process,
along with the tag specified in the \texttt{spawn} statement.
In Figure~\ref{fig:inc}, the created processes have name tags
\texttt{process/0} and \texttt{process/1}.  The main process that
starts the execution has name tag \texttt{\_\_main\_\_/()}.

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty tuple, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
(In CXL, dictionary values are enclosed by \texttt{dict\{} and \texttt{\}}.)
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

\chapter{CXL Methods and Dictionaries}

CXL methods look and feel much the same as Python methods, but there
are some differences you should be aware of.
First, as already noted, there is no \texttt{return} statement.
A CXL method returns when the last statement in the method has
completed.

Like in Python, CXL methods have local variables.  You have already
seen two types of local variables.
First, the arguments to the method are variables, although read-only.
Second, the variable \texttt{result} can be assigned to
change the default return value of a method, which is \texttt{()}.
You can declare additional local variables to hold intermediate
results.  For example, you can
declare a local variable $x$ with initial value 3 as follows:
\begin{code}
\begin{verbatim}
var x = 3;
\end{verbatim}
\end{code}
The initial value is mandatory.
Unless a variable is declared this way, it will be considered
a global shared variable.
The local variables of a method cannot be accessed other than by
the method itself.

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{x}).
\begin{code}
\begin{verbatim}
var x = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.  If you invoke
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.

If you want to invoke a method but are not interested in its result,
you have to use a \texttt{call} statement like so:
\begin{code}
\begin{verbatim}
call m a;
\end{verbatim}
\end{code}

\chapter{Spinlock}
Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.
